Bayesian Sparsification of $\cplx$-valued networks

### 2019-11-08 paper/paper.tex
listed properties of Ei (with proof) compressed improper prior section

In this section we consider two different priors, suitable for variational dropout: log-uniform
\cite{kingma_variational_2015,molchanov_variational_2017} and proper $\cplx$-Gaussian analogue
for the empirical Bayes prior in \cite{kharitonov_variational_2018}.

Recall that we use fully factorized variational approximation for the parameters: $
  W \sim q(W)
    = \prod_{ij} q(\omega_{ij})
$ with proper complex Gaussian distributions of each element $
  q(\omega_{ij})
    = \mathcal{N}^{\cplx}\bigl(
      \omega_{ij} \big\vert
      \mu_{ij}, \sigma^2_{ij}, 0
    \bigr)
$. Assuming factorized prior belief $\pi(W)$ on the parameters $W$, the KL-divergence
term in ELBO \eqref{eq:elbo} is
\begin{equation}  \label{eq:elbo-general-kl-div}
KL(q(W) \| \pi(W))
  % = \mathbb{E}_{W \sim q(W)} \log \tfrac{q(W)}{\pi(W)}
  % = \sum_{ij} \mathbb{E}_{\omega_{ij} \sim q(\omega_{ij})}
  %   \log \tfrac{q(\omega_{ij})}{\pi(\omega_{ij})}
  = \sum_{ij} KL\bigl(
    q(\omega_{ij}) \| \pi(\omega_{ij})
  \bigr)
  \,.
\end{equation}
%
We shall omit the subscripts for brevity in the derivations in the next section, which
we begin with a derivation of the entropy of a proper $\cplx$-Gaussian.

The expression of Kullback-Leibler divergence of $q(\omega)$ from an arbitrary prior $\pi(\omega)$
necessarily involves the (differential) entropy of $q$:
\begin{equation*}
  KL\bigl(
    q(\omega) \| \pi(\omega)
  \bigr)
    % = \mathbb{E}_{\omega \sim q} \log{q(\omega)} - \mathbb{E}_{\omega \sim q} \log{\pi(\omega)}
    = - \mathbb{H}(q) - \mathbb{E}_{\omega \sim q} \log{\pi(\omega)}
    \,.
\end{equation*}

@@

Let's get the divergence term for the improper prior, that most closely recovers the variational
dropout penalty term in \cite{kingma_variational_2015} and \cite{molchanov_variational_2017}.
If we assume $
  \pi(\omega) \propto \tfrac1{\lvert \omega \rvert^\beta}
$ for some $\beta > 1$, then we formally get the following KL-divergence term (up to an unknown,
or non-existent constant):
\begin{equation}  \label{eq:var-do-kl-div-raw}
KL(q\| \pi)
  \propto
  %   \mathbb{E}_{\omega \sim q(\omega)} \log q(\omega)
  %   + \tfrac{\beta}2 \mathbb{E}_{\omega \sim q(\omega)} \log \lvert \omega \rvert^2
  % =
  %   - \log \lvert \det{(\pi e \sigma^2)} \rvert
  %   + \tfrac{\beta}2 \mathbb{E}_{\omega \sim q(\omega)} \log \lvert \omega \rvert^2
  % =
    - \log{\sigma^2}
    + \tfrac{\beta}2 \mathbb{E}_{\omega \sim q(\omega)} \log \lvert \omega \rvert^2
  \,,
\end{equation}
where we have used \eqref{eq:cn-proper-entropy} with $\Gamma = \sigma^2 \geq 0$. Note
that for $\mu \neq 0$, we have $\omega \sim q(\omega)$ equal in distribution to $\mu \xi$
with $\xi \sim \mathcal{N}^{\cplx}(\xi \vert 1, \alpha, 0)$, whence
\begin{equation}  \label{eq:expect-improper-term-cplx}
\mathbb{E}_{\omega \sim q(\omega)} \log \lvert \omega \rvert^2
  % = \mathbb{E}_{\xi \sim \mathcal{N}^{\cplx}(\xi \vert 1, \alpha, 0)}
  %   \log \lvert \mu \xi \rvert^2
  % = \log \lvert \mu \rvert^2
  %   + \mathbb{E}_{\xi \sim \mathcal{N}^{\cplx}(\xi \vert 1, \alpha, 0)}
  %       \log \lvert \xi \rvert^2
  = \log \alpha \lvert \mu \rvert^2
    + \mathbb{E}_{\xi \sim \mathcal{N}^{\cplx}(\xi \vert 0, 1, 0)}
        \log{\bigl\lvert \tfrac1{\sqrt{\alpha}} + \xi \bigr\rvert^2}
  \,.
\end{equation}
The expectation in \eqref{eq:expect-improper-term-cplx} involves a log-moment of the non-central
$\chi^2_2$ distribution, which has an analytic expression for the gradient, unlike the case
in $\real$-dropout penalty \eqref{eq:improper-kl-div-real}.

Let $(\mu_i)_{i=1}^m \in \cplx$ and $
  (z_i)_{i=1}^m \sim \mathcal{N}^{\cplx}(0, 1, 0)
$ iid. Then $W = \sum_i \lvert \theta_i + z_i \rvert^2$ has a non-central $\chi^2_{2m}$
distribution with non-centrality parameter $s^2 = \sum_i \lvert \theta_i \rvert^2$. The
log-moments of $W$ for general integer $m \geq1$ are given in
\cite[p.~2466]{lapidoth_capacity_2003}.  % in fact Appendix X, Lemma 10.1
In particular, for $m=1$ and $\theta\in \cplx$ we have
\begin{equation}  \label{eq:log-moment-for-chi-2}
\mathbb{E}_{z \sim \mathcal{N}^{\cplx}(0, 1, 0)}
  \log \lvert \theta + z \rvert^2
  = \log \lvert \theta \rvert^2 - \mathop{Ei}( - \lvert \theta \rvert^2)
  \,,
\end{equation}
where $\mathop{Ei}(x) = \int^x_{-\infty} t^{-1} e^t dt$ for $x < 0$ is the exponential integral
function. Note that the limiting value of \eqref{eq:log-moment-for-chi-2} as $
  \lvert \theta \rvert \to 0
$ is $- \gamma$, i.e. negative Euler's constant. Therefore after cancellation the expression
of the KL-divergence in \eqref{eq:var-do-kl-div-raw} becomes
\begin{equation}  \label{eq:var-do-kl-div}
KL(q\| \pi)
  \propto
  %   - \log \lvert \det{(\pi e \sigma^2)} \rvert
  %   + \tfrac{\beta}2 \log \alpha \lvert \mu \rvert^2
  %   + \tfrac{\beta}2 \mathbb{E}_{\xi \sim \mathcal{N}^{\cplx}(\xi \vert 0, 1, 0)}
  %       \log{\bigl\lvert \tfrac1{\sqrt{\alpha}} + \xi \bigr\rvert^2}
  % =
  %   - \log{\sigma^2}  % - \log{\pi e}
  %   + \tfrac{\beta}2 \log \alpha \lvert \mu \rvert^2
  %   + \tfrac{\beta}2 \bigl(
  %     - \log \alpha - \mathop{Ei}( -\tfrac1{\alpha})
  %   \bigr)
  % =
  %   - \log{\sigma^2}
  %   + \tfrac{\beta}2 \log \alpha \lvert \mu \rvert^2
  %   - \tfrac{\beta}2 \log \alpha
  %   - \tfrac{\beta}2 \mathop{Ei}( -\tfrac1{\alpha})
  % =
    % \log \tfrac{\lvert \mu \rvert^\beta}{\sigma^2}
    % \log \tfrac{\lvert \mu \rvert^\beta}{\alpha \lvert \mu \rvert^2}
    \tfrac{\beta-2}2 \log{\lvert \mu \rvert^2}
    + \log{\tfrac1{\alpha}}
    - \tfrac{\beta}2 \mathop{Ei}(- \tfrac1{\alpha})
  \,.
\end{equation}

The exponential integral is readily available in most scientific computations packages nowadays,
and, what is especially useful for the stochastic variational Bayes purposes, has a simple
expression for the derivative: $x \mapsto \tfrac{e^x}{x}$ for $x < 0$. This observation enables
embedding it into autograd software by computing the term with bogus output values, yet correct
(on device) gradient, provided no downstream computations depend on the value of the KL-divergence
term itself.

@@

>>>>

We consider two priors for Bayesian dropout: log-uniform and $\cplx$-Gaussian empirical Bayes,
\cite{kingma_variational_2015,molchanov_variational_2017,kharitonov_variational_2018}. For a 
fully factorized approximation $q(\omega)$ and factorized prior belief $\pi(\omega)$, the
divergence term in ELBO \eqref{eq:elbo} is
\begin{equation}  \label{eq:elbo-general-kl-div}
KL(q(\omega) \| \pi(\omega))
  % = \mathbb{E}_{\oemga \sim q(\oemga)} \log \tfrac{q(\oemga)}{\pi(\oemga)}
  % = \sum_{ij} \mathbb{E}_{\omega_{ij} \sim q(\omega_{ij})}
  %   \log \tfrac{q(\omega_{ij})}{\pi(\omega_{ij})}
  % = \sum_{ij} KL\bigl( q(\omega_{ij}) \| \pi(\omega_{ij}) \bigr)
  = - \sum_{ij} \mathbb{H}(q(\omega_{ij}))
    + \mathbb{E}_{\omega_{ij} \sim q(\omega_{ij})} \log{\pi(\omega_{ij})}
  \,.
\end{equation}
We use approximation \eqref{eq:c-gauss-vi} and omit subscripts ${ij}$ below for brevity.

@@

From \eqref{eq:cn-proper-entropy} the KL-divergence for \verify{log-uniform prior} $
  \pi(\omega) \propto {\lvert \omega \rvert}^{-\beta}
$ with $\beta \geq 1$ is
\begin{equation}  \label{eq:var-do-kl-div-raw}
KL(q\| \pi)
  \propto
  %   \mathbb{E}_{\omega \sim q(\omega)} \log q(\omega)
  %   + \tfrac{\beta}2 \mathbb{E}_{\omega \sim q(\omega)} \log \lvert \omega \rvert^2
  % =
  %   - \log \lvert \det{(\pi e \sigma^2)} \rvert
  %   + \tfrac{\beta}2 \mathbb{E}_{\omega \sim q(\omega)} \log \lvert \omega \rvert^2
  % =
    - \log{\sigma^2}
    + \tfrac{\beta}2 \mathbb{E}_{\omega \sim q(\omega)} \log \lvert \omega \rvert^2
Note that for $\mu \neq 0$ from \eqref{eq:cn-affine} we have $
  \mathcal{N}^{\cplx}(\mu, \alpha \lvert \mu \rvert^2, 0)
  \sim \mu \cdot \mathcal{N}^{\cplx}(1, \alpha, 0)
$, whence
\begin{equation}  \label{eq:expect-improper-term-cplx}
\mathbb{E}_{\omega \sim q(\omega)} \log \lvert \omega \rvert^2
  % = \mathbb{E}_{\xi \sim \mathcal{N}^{\cplx}(\xi \vert 1, \alpha, 0)}
  %   \log \lvert \mu \xi \rvert^2
  % = \log \lvert \mu \rvert^2
  %   + \mathbb{E}_{\xi \sim \mathcal{N}^{\cplx}(\xi \vert 1, \alpha, 0)}
  %       \log \lvert \xi \rvert^2
  = \log \alpha \lvert \mu \rvert^2
    + \mathbb{E}_{\xi \sim \mathcal{N}^{\cplx}(\xi \vert 0, 1, 0)}
        \log{\bigl\lvert \tfrac1{\sqrt{\alpha}} + \xi \bigr\rvert^2}
Let $(\mu_i)_{i=1}^m \in \cplx$ and $
  (z_i)_{i=1}^m \sim \mathcal{N}^{\cplx}(0, 1, 0)
$ iid. Then $W = \sum_i \lvert \theta_i + z_i \rvert^2$ has a non-central $\chi^2_{2m}$
distribution with non-centrality parameter $s^2 = \sum_i \lvert \theta_i \rvert^2$. The
log-moments of $W$ for general integer $m \geq1$ are given in
\cite[p.~2466]{lapidoth_capacity_2003}.  % in fact Appendix X, Lemma 10.1
In particular, for $m=1$ and $\theta\in \cplx$ we have
\begin{equation}  \label{eq:log-moment-for-chi-2}
\mathbb{E}_{z \sim \mathcal{N}^{\cplx}(0, 1, 0)}
  \log \lvert \theta + z \rvert^2
  = \log \lvert \theta \rvert^2 - \mathop{Ei}( - \lvert \theta \rvert^2)
where $
  \mathop{Ei}(x) = \int^x_{-\infty} t^{-1} e^t dt
$ for $x < 0$ is the Exponential Integral. Note that
$
  \tfrac{d}{dx} \mathop{Ei}(x) = \tfrac{e^x}{x}
  \mathop{Ei}(x) \leq \log{(-x)}
$ on $x < 0$, asymptotically $
  \log{(-x)} - \gamma
$ as $x\to 0-$ and $
  \mathop{Ei}(x) \geq -e^x
$ on $x \in (-\infty, -1]$ ($\gamma$ is Euler's constant).
% 1. by Leibniz rule $Ei(x)$ has -ve derivative on $x < 0$
% 2. $Ei(x) \leq 0$ on $x < 0$, since $t^{-1} e^t \leq t^{-1} < 0$ on $t < 0$
% 3. $Ei(x) \leq \log{(-x)}$ on $x < 0$, since on $x \in [-1, 0)$
%   \mathop{Ei}(x)
%     = \mathop{Ei}(-1) + \int_{-1}^x t^{-1} e^t dt
%     \leq \mathop{Ei}(-1) + \int_{-1}^x t^{-1} dt
%     = \mathop{Ei}(-1) + \log{(-x)}
%     \leq \log{(-x)}
% 3. by l'H{\^o}pital rule $Ei(x)$ is asymptotically $\log{-x}$ as $x \to 0-$
% % t = -e^{-x} ,  -\log{(-t)} = x
%   \lim_{x\to 0-} \frac{\mathop{Ei}(x)}{\log{(-x)}}
%     = \lim_{x\to 0-} \frac{x^{-1} e^x}{x^{-1}}
%     = 1
% 4. \cite{lapidoth_capacity_2003} yields $\lim_{x\to 0-} \log{(-x)} - \mathop{Ei}(x) = \gamma$
% 5. $Ei(x) \geq -e^x$ for any $x < -1$, since $t^{-1} e^t \geq - e^t$ for $t < -1$
% \mathop{Ei}(x)
%   = \int_{-\infty}^x t^{-1} e^t dt
%   \geq \int_{-\infty}^x -e^t dt = -e^x
% 6. $Ei(x) \leq \log{(-x)} - \gamma$ on $x < 0$ and $-e^x \leq Ei(x) \leq 0$ on $x < -1$.
After cancellation \eqref{eq:var-do-kl-div-raw} becomes
\begin{equation}  \label{eq:var-do-kl-div}
KL(q\| \pi)
  \propto
  %   - \log \lvert \det{(\pi e \sigma^2)} \rvert
  %   + \tfrac{\beta}2 \log \alpha \lvert \mu \rvert^2
  %   + \tfrac{\beta}2 \mathbb{E}_{\xi \sim \mathcal{N}^{\cplx}(\xi \vert 0, 1, 0)}
  %       \log{\bigl\lvert \tfrac1{\sqrt{\alpha}} + \xi \bigr\rvert^2}
  % =
  %   - \log{\sigma^2}  % - \log{\pi e}
  %   + \tfrac{\beta}2 \log \alpha \lvert \mu \rvert^2
  %   + \tfrac{\beta}2 \bigl(
  %     - \log \alpha - \mathop{Ei}( -\tfrac1{\alpha})
  % =
  %   - \log{\sigma^2}
  %   + \tfrac{\beta}2 \log \alpha \lvert \mu \rvert^2
  %   - \tfrac{\beta}2 \log \alpha
  %   - \tfrac{\beta}2 \mathop{Ei}( -\tfrac1{\alpha})
  % =
    % \log \tfrac{\lvert \mu \rvert^\beta}{\sigma^2}
    % \log \tfrac{\lvert \mu \rvert^\beta}{\alpha \lvert \mu \rvert^2}
    \tfrac{\beta-2}2 \log{\lvert \mu \rvert^2}
    + \log{\tfrac1{\alpha}}
    - \tfrac{\beta}2 \mathop{Ei}(- \tfrac1{\alpha})
% For $\beta = 2$ we add $\gamma$ to make the right-hand side nonegative. (dubious)
Gradient-based methods that optimize \eqref{eq:elbo} usually do not rely on the value
of the objective. Since $\mathop{Ei}$ has simple analytic derivative and downstream
computations depend \eqref{eq:var-do-kl-div} only additively in\eqref{eq:elbo}, it is
possible to backpropagate through the divergence without forward evaluation.

@@

The expectation in \eqref{eq:expect-improper-term-cplx} involves a log-moment of the non-central
$\chi^2_2$ distribution, which has an analytic expression for the gradient, unlike the case
in $\real$-dropout penalty \eqref{eq:improper-kl-div-real}.
